\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
Incorporate two comments of Andy and Kyle

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

For hyperbolic problems, we integrate the solution on a a single uniform
patch, containing $m^d$ cells,
using the wave propagation algorithm described by R. J. LeVeque and
implemented in \clawpack \cite{le:2002, clawpack}.  We assume a
single degree of freedom per cell and reconstruct a piecewise
constant solution to obtain left and right states at cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface.  For non-linear
problems and systems, an approximate Riemann solver, such as a Roe solver
\cite{roesolver}, is typically used.  Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

>>>>

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
% CB: dropped dependencies.
% dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

For hyperbolic problems, we integrate the solution on a a single uniform
patch, containing $m^d$ cells,
using the wave propagation algorithm described by R. J. LeVeque and
implemented in \clawpack \cite{le:2002, clawpack}.  Optionally, the coordinates
of the patch can be geometrically mapped.  We assume a
single degree of freedom per cell and reconstruct a piecewise
constant solution to obtain left and right states at cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface.  For non-linear
problems and systems, an approximate Riemann solver, such as a Roe solver
\cite{roesolver}, is typically used.  Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.
